SSA based global code motion of identical computations

### 2017-04-14 hoist.tex
remove dfsin formalisms

The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Single Entry Single Exit (\SESE{}) \cite{sese} and Single Entry Multiple Exit
  (SEME) regions;
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

@@

\subsection{Finding candidates to hoist}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}.

The current implementation of \GVN{} in \LLVM{} has some limitations when it comes to
loads and stores so we compute the \GVN{} of loads and stores separately.  Our
short-time solution to value number loads is to hash the address from where the
value is to be loaded. For stores, we value number the address as well as the
value to be stored at that address.

@@

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby in cases when
there are no fully redundant instructions. Essentially if,

\begin{verbatim}
// B1 != B2 != B3
BasicBlock B1, B2, B3
DFSIn(B1) = depth first discovery time stamp
DFSIn(B1,B2) = |DFSIn(B1) - DFSIn(B2)|
DFSIn(B1,B3) = |DFSIn(B1) - DFSIn(B3)|

Depth(B1) = depth of tree from the root node.
// When B1 dominates B2
BBDist(B1, B2) = |Depth(B1) - Depth(B2)|

NCD(B1, B2) = nearest common dominator of B1 and B2
BBDist(B1,B2) = BBDist(B1, NCD(B1, B2))
BBDist(B1,B3) = BBDist(B1, NCD(B1, B3))

Then,
DFSIn(B1,B2) < DFSIn(B1,B3) implies
 BBDist(B1,B2) <= BBDist(B1,B3)
\end{verbatim}

Same property would hold for instructions I1, I2, I3 provided they are not fully
redundant, i.e., none of the basic blocks containing I1, I2 and I3 respectively
dominate the other:

\begin{verbatim}
Instruction I1, I2, I3
// I1, I2, I3 are not fully redundant
// B1, B2, B3 contains I1, I2, I3 respectively
DFSIn(I1) = DFSIn(B1)
BBDist(I1, I2) = BBDist(B1, B2)
DFSIn(I1) = DFSIn(B1)
DFSIn(I1,I3) = DFSIn(B1, B3)
NCD(I1, I2) = NCD(B1, B2)

Then,
 DFSIn(I1,I2) < DFSIn(I1,I3) implies
  BBDist(I1,I2) <= BBDist(I1,I3)
\end{verbatim}

It says that if I1 is closer to I2 than I3 in the list of candidates sorted by
DFSIn numbers, then the nearest common dominator of I1 and I2 will be closer (in
terms of basic block distance) to I1, than nearest common dominator of I1 and
I3.

In the presence of fully redundant computations in the list of hoistable
candidates this equation may not hold. For example, if there is another
instruction I4 which is dominated by I1, but I4 is present in a basic block
which is farther than $DFSIn(I1, I2)$, i.e., $DFSIn(I1,I2) < DFSIn(I1,I4)$;
however, $BBDist(I1,I4) = 0 < BBDist(I1,I2) >=1$.

>>>>

The algorithm for code hoisting uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Control Flow Graph (\CFG{}) and the Dominance (DOM) and Post-Dominance (PDOM)
  relations \cite{dragonbook};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given a
  unique number and the expressions that the compiler can prove identical are
  given the same number;
\item Memory \SSA{} \cite{novillo2007memory}: memory operations that the compiler
  is able to prove in dependence are linked through use-def chains.
\end{itemize}

@@

\subsection{Finding candidates to move}
The first step is to find a set of instructions that perform identical
computations: this is performed by a linear scan of all instructions of the
program and classifying all instructions by their value given by \GVN{}.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution to value number loads is to hash the address from
where the value is to be loaded. For stores, we value number the address as well
as the value to be stored at that address.

@@

\subsubsection{Partition the list of hoisting candidates to maximize hoisting}
\label{subsec:partition}
In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoistings.  By sorting
the list of all the candidates in the increasing order of their depth first
search discovery time stamp \cite{clrs} (DFSIn numbers), we make sure that
candidates closer in the list have their common dominator nearby in cases when
there are no fully redundant instructions. In Figure-\ref{fig:dfsin}, if B3 B4
B5 and B6 have identical computations and for some reason they cannot be hoisted
together at B0, then we partition the set of hoistable candidates in their DFSin
order.

\begin{figure}
\centering
%rankdir=LR;
\digraph[scale=0.35]{abc}{
  B0 -> B1
  B0 -> B2
  B1 -> B3
  B1 -> B4
  B2 -> B5
  B2 -> B6
}
\caption{CFG to illustrate partitioning}
\label{fig:dfsin}
\end{figure}
